Informative Autonomous Sampling of Oceanographic Variables Using Joint Excursion Sets

### 2019-10-09 main_new.tex
Small updates

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand dynamic processes in the water-column, the
  combination of statistics and autonomous robotics provide
  opportunities for studying experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal processes by providing informative descriptions of spatial
  characteristics of ocean phenomena. Specifically, we define a design
  criterion based on reducing uncertainty in the joint excursions of
  random fields, and derive analytical expressions for the expected
  Bernoulli variance reduction in the excursion set of spatial
  temperature and salinity variables. We show how these can be
  applicable in static and sequential designs, building on myopic and
  non-myopic sampling strategies. We use simulations to study the
  properties and differences between state-of-the-art approaches
  followed by results from field deployments with an autonomous
  underwater vehicle as part of a case study mapping the boundary of a
  river plume. The results demonstrate the potential of combining statistical methods and robotic platforms, and the abilities of the derived methods to effectively inform and execute data-driven environmental sampling.

@@

Traditional data collection at sea has typically been based on static
buoys, Lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean and are impacted by cloud cover. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have
advanced the state of sampling and consequently have made robotics an
integral part of ocean observation; our previous work has contributed
to this effort
% using statistical methods in design and data assimilation
\citep{das11b,Das2015,fossum18b,fossuminformation}.  Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.

The potential that lies in using information-based methods, which we
highlight in the paper, is to do efficient environmental sampling and
overcome three common challenges, namely:

@@

\item As a priori information about the state of the ocean is
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken in-situ. Recently
  acquired information can be assimilated into statistical models that
  can be used to inform decisions on where to sample
  sequentially.%\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.

\item For sampling problems related to environmental sensing, the
  number of choices (i.e. locations, trajectories, and candidate
  designs) makes the problem combinatorially large, creating a
  trade-off between optimization (finding the most resource-efficient
  design to collect necessary data), and computability (arriving at a
  solution in reasonable time). To successfully resolve features, this
  trade-off has to be considered in sample design and the practice.

@@

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. In our formulation we use
Gaussian Processes (GPs) in large part because they allow for
efficient implementation and evaluation onboard a robotic platform.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography, such as fresh water inflows) and
incipient micro-biology driven by planktonic and coastal anthropogenic
input, such as pollution and agricultural runoff. These often result
in a range of ecosystem-related phenomena such as blooms, anoxic zones
and plumes, all of which have significant direct and indirect effects
on society. Walter Munk referred to the 20\textsuperscript{th} century
as the \emph{century of undersampling} \citep{munk2002}, pointing out
a lack of sufficient sampling resolution in time and space as a
critical issue for characterizing ocean phenomena. Even today,
policymakers struggle with limited observability of severe
environmental problems; a prominent example is the recent (May 2018)
outbreak of harmful algal blooms in Northern Norway\footnote{Report
  from the Institute of Marine Research: https://bit.ly/2KZGJfI}, with
losses exceeding 12 million tons of salmon valued at \$80 million
(Norwegian Directorate of Fisheries). This is a case in point of how
adequate sampling efforts could have provided a more accurate
assessment and early warning of changing environmental
conditions. These challenges related to undersampling have been
driving ocean sensing practices towards more autonomous, networked,
and distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir1}), where observation is guided
by assimilation of measurements and statistical models.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $5 m^2$ -- $10 km^2$) where
lateral spatial variability tends to dominate. Remote sensing and
synthetic ocean models often find it challenging to provide detail at
this scale \citep{Lermusiaux:2006} and the principal way to resolve
for fine resolution is by direct in-situ observation. At larger
\emph{mesoscale} ($>50 km^2$), both three-dimensional space and time
dynamics are important and can shift substantially, while in the case
of river plumes one can often limit the scope to cover only lateral
spatial elements. However, the methodological framework can be
extended to higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity dependent on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources, one can augment information from synthetic ocean models,
remote-sensing data, or fixed-location sensors on buoys. Traditional
operations and surveys with AUVs are usually limited to observations
along fixed transects on a spatial grid, pre-programmed by the human
operator. Instead using onboard algorithms to continuously evaluate,
update, and refine future sampling locations, one can make information
gathering \emph{adaptive} and \emph{contextual} to the survey
requirements \citep{das11b,fossum18b,fossuminformation}. The space of
sampling opportunities is still limited to a grid, but adaptation
allows more flexibility. For instance, an AUV could reconstruct or
modify a survey line based on what temperatures it is measuring by
using embedded methods in decision-theoretic planning
\citep{py10,Rajan12,Rajan12b}.

@@

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:envir1}). Myopic sampling, i.e. step-wise selection of
the path (on the graph) which reduces the information criterion the
most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward the
designs that improve the classification of water masses as a means to
characterize the frontal zone.

@@

We consider a lateral domain near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes:
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. The generalization
to more than two variables is given in Section \ref{gen_k}. The
adaptive sampling strategies are described in Section
\ref{sec:heuristics}.

@@

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}) and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense,
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
the IBV for univariate processes. A major contribution of our work is
to derive closed-form results for this design criteria for situations
where the underlying model is based on multivariate GPs.

@@

We discretize the spatial domain $\mathcal{M}$ to a set of $n$ grid
locations $\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell
has area $\Delta$; the grid is used for the waypoint graph setting the
possible design locations. It is also used when solving integral
expressions in Eq. (\ref{measVAR}) and (\ref{sur}), which are then
approximated by sums over all grid cells.  Using vector notation, we
denote the Gaussian temperature and salinity variables on the grid and
its density function by the formula:
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma). %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

An important concept underlying this work is the potential of robotic
sampling to adapt survey plans based on data and statistical methods
in accordance with the sense-plan-act loop (Fig. \ref{fig:envir1}). We
will therefore present a few adaptive strategies for robotic sampling
that build on the notion of multivariate ES.

@@

An adaptive AUV survey is split into many rounds (stages). The selection is made on the graph and defined by the nearest grid
nodes in the domain $\mathcal{M}_g$.

The optimal sequential design not only considers the best current AUV
grid node, but also how the data gathered at this node could inform
future sampling at successive stages. Let $d^{j,s}$ denote design
number $j$ at stage $s$ of a sequential survey. If this design is
selected, data $\by^{j,s}$ will be gathered. The optimal path
selection situation is depicted in Fig. \ref{fig:PathSelOpt},
\begin{figure}[b!]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a
series of intermixed maximizations over designs and integrals over
data.  In practice, the optimal solution is intractable because of
the enormous growth over stages (see
e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}).
Instead, we outline heuristic strategies. In terms of notation, let
$\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote
the data gathered until stage $s-1$, using a selected design
$d^{j,r}$, $r=1,\ldots,s-1$.

@@

\subsection{Myopic Path Planning}
\label{myopic}

@@

\subsection{Look-ahead Trajectory Planning}
\label{LA}

@@

We use a bivariate GP model for temperature and salinity, where we
specify the prior mean:
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{T,0} + \beta_{T,1} x_{\mbox{West}} \\
\beta_{T,1} + \beta_{S,1} x_{\mbox{West}}
\end{bmatrix}.
\end{equation}
This situation
mimics that of a river mouth opening from the south, with the water
masses pulled to the east. The regression parameters are  $\beta_{T,1}=0.065$ and
$\beta_{S,1}=0.1$ for the slope, with west coordinate and $\beta_{T,0}=5.8$
and $\beta_{S,1}=29.0$ for the intercept at a reference start map coordinate.

@@

We compare three different static designs denoted
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} with the three sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. Recall that
the static sampling paths are pre-determined and cannot be altered.
These represent the pre-planned strategies used in most current AUV
operational survey designs.

@@

The tables below explore the sensitivity of the result to different
prior correlation values between temperature measured onboard the AUV,
temperature and salinity, prior standard deviations, and spatial
correlations (see table rows). We only compare results at the end of
each survey. The expected ES variances are shown in Table
\ref{tab:sim_res_ev}, results of RMSE are in Table
\ref{tab:sim_res_rmse}, and explained variance is in Table
\ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and
\textit{look-ahead} strategies perform the best in terms of expected
variance in the ES. Surprisingly, the \textit{myopic} strategy
performs slightly better than the \textit{look-ahead}. Part of the
reason for this discrepancy can be traced to the act of pruning some
of the paths for the \textit{look-ahead} strategy; however, even with
less pruning, the \textit{look-ahead} does not outperform the
\textit{myopic} approach. The \textit{static\_north} design is clearly
better than the other two static designs, the reason being that the
others do not focus on the areas where the ES is really uncertain
(centre locations are the most uncertain). By measuring temperature
alone, the expected variances in the ES are only slightly higher than
the combined temperature-salinity estimate. This occurs because the
correlation of $0.6$ means that temperature observations provide
information about salinity as well. For the other predictive measures,
\textit{myopic} and \textit{look-ahead} strategies do well, clearly
beating the \textit{naive} strategy. The \textit{static\_zigzag}
design is usually the best among the static designs for these
predictive performance measures because its pattern gives more
effective coverage of the domain when there is significant spatial
correlation, as indicated in this case.

@@

\subsection{Model Specification}

The statistical model parameters were specified based on a short
preliminary survey where the AUV made an initial transect to determine
environmental conditions and correlation structure. Based on the
initial data, the trend parameters were estimated by linear
regression, where both temperature and salinity are assumed to
increase linearly, going west from the river mouth. Next, the
residuals from the regression analysis were analyzed to study the fit
of the GP model and to specify the covariance parameters.

@@

Fig. \ref{fig:parest} shows diagnostic plots of this
analysis. Fig. \ref{fig:parest_a} shows a cross-plot of temperature
and salinity residuals after the westerly trends in both salinity and
temperature are subtracted from the data. This scatterplot of joint
residuals indicates larger variability in salinity than in
temperature, and a positive correlation ($0.5$) between the two
variables. Based on the fitted bivariate Gaussian model (ellipses in
Fig. \ref{fig:parest_a}), we can compute the modeled quadratic form of
the residuals, and if the model is adequate they should be
approximately $\chi^2_2$ distributed. Fig. \ref{fig:parest_b} shows
the empirical cumulative distribution function (CDF) of these
quadratic forms (solid) together with the theoretical CDF of the
$\chi^2_2$ distribution. The modeled and theoretical curves are very
similar, which indicates that the Gaussian model fits reasonably
well. Even though there appears to be some clustering in both
Fig. \ref{fig:parest_a} and \ref{fig:parest_b}, the bivariate
diagnostic plots look reasonable and justify a Gaussian
model. Fig. \ref{fig:parest_c} shows the empirical variogram of the
scaled residuals for temperature and salinity. The decay is similar
for the two, and seems to be negligible after about $150$ m.

@@

The sampling locations were distributed over an equilateral grid, as
shown in the grey-colored lattice in Fig. \ref{fig:map}. The robotic
platform consisted of a Light AUV \citep{sousa2012lauv}
(Fig. \ref{fig:lauv}) equipped with a 16 Hz Seabird Fastcat-49
conductivity, temperature, and depth (CTD) sensor providing
temperature and salinity
measurements. %The accuracy of the CTD instrument is $\pm 0.0003$ S/m (conductivity) and $\pm0.002\,^{\circ}\mathrm{C}$ (temperature).
The sampling agent was built on top of the autonomous agent framework
Teleo-Reactive EXecutive (\textit{T-REX})
\citep{py10,Rajan12,Rajan12b}, running an instance of the
\textit{myopic} strategy from Section \ref{sec:sampling_designs} to
control the AUV and decide between sampling locations.

@@

The sampling strategy was designed around the concept of visiting
waypoints sequentially. Arriving at a desired waypoint with new
measurements and an updated model, the AUV triggers the myopic
strategy to evaluate all possible design criteria (see
Eq. \eqref{critSEQ}). The waypoint-and-path combination that is
expected to reduce the IBV the most is selected; upon arrival this
procedure is then repeated. At each stage, it takes the AUV about 30
seconds to evaluate the expected IBV for all the possible
waypoint-and-path alternatives.

The AUV was set to start in the south-center part of the waypoint
graph, with the previously outlined GP model of the environment. A
survey was set to take approximately 40 minutes to visit 15 waypoints
on the grid, with the AUV running near the surface to capture the
plume. On its path from one waypoint to the next, the AUV gathered
data regularly, and the GP model assimilated temperature and salinity
data with an update frequency of 30 seconds, with about three updates
per stage.

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand dynamic processes in the water-column, the
  combination of statistics and autonomous robotics provide new
  opportunities for studying experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal processes by providing informative descriptions of spatial
  characteristics of ocean phenomena. Specifically, we define a design
  criterion based on reducing uncertainty in the joint excursions of
  random fields, and derive analytical expressions for the expected
  Bernoulli variance reduction in the excursion set of spatial
  temperature and salinity variables, providing a description of
the boundary uncertainty that can be used prioritize sampling efforts at locations that are
ambiguous, making exploration more effective. We show how these can be
  applicable in static and sequential designs, building on myopic and
  non-myopic sampling strategies. We use simulations to study the
  properties and differences between state-of-the-art approaches,
  followed by results from field deployments with an autonomous
  underwater vehicle as part of a case study mapping the boundary of a
  river plume. The results demonstrate the potential of combining statistical methods and robotic platforms, and the abilities of the derived methods to effectively inform and execute data-driven environmental sampling.

@@

Traditional data collection at sea has typically been based on static
buoys, Lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have
advanced the state of sampling and consequently have made robotics an
integral part of ocean observation; our previous work has contributed
to this effort \citep{das11b,Das2015,fossum18b,fossuminformation}. Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.

The potential that lies in using information-based methods, which we
highlight in this paper, is to do efficient environmental sampling and
overcome three common challenges, namely:

@@

\item As a priori information about the state of the ocean is
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken in-situ. Recently
  acquired information can be used to resolve this by assimilating up-to-date data into statistical models that
  can be used to inform decisions on where to sample
  sequentially.%\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.

\item For sampling problems related to environmental sensing, the
  number of choices (i.e. locations, trajectories, and candidate
  designs) makes the problem combinatorially large, creating a
  trade-off between optimization (finding the most resource-efficient
  design to collect necessary data), and computability (arriving at a
  solution in reasonable time). To successfully resolve features, this
  trade-off has to be considered in development and practice.

@@

Addressing this, the combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. In our formulation we use
Gaussian Processes (GPs) in large part because they allow for
efficient implementation and evaluation in real time onboard a robotic platform.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences, and local geography, such as fresh water inflows) and
incipient micro-biology driven by planktonic and coastal anthropogenic
input, such as pollution and agricultural runoff transported into the ocean by rivers and streams. These often result
in a range of ecosystem-related phenomena such as blooms, anoxic zones
and plumes, all of which have significant direct and indirect effects
on society. Walter Munk referred to the 20\textsuperscript{th} century
as the \emph{century of undersampling} \citep{munk2002}, pointing out
a lack of sufficient sampling resolution in time and space as a
critical issue for characterizing ocean phenomena. Even today,
policymakers struggle with limited observability of severe
environmental problems; a prominent example is the recent (May 2018)
outbreak of harmful algal blooms in Northern Norway\footnote{Report
  from the Institute of Marine Research: https://bit.ly/2KZGJfI}, with
losses exceeding 12 million tons of salmon valued at \$80 million
(Norwegian Directorate of Fisheries). This is a case in point of how
adequate sampling efforts could have provided a more accurate
assessment and early warning of changing environmental
conditions. These challenges related to undersampling have been
driving ocean sensing practices towards more autonomous, networked,
and distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir1}), where observation is guided
by assimilation of recent measurements and statistical models.

@@

River plumes belong to a class of ocean processes that are local and
act on smaller \emph{sub-mesoscale} extent (from $5 m^2$ -- $10 km^2$) where
lateral spatial variability tends to dominate. Remote sensing and
synthetic ocean models often find it challenging to provide detail at
this scale \citep{Lermusiaux:2006} and the principal way to resolve
for fine resolution is by direct in-situ observation. At larger
\emph{mesoscale} ($>50 km^2$), both three-dimensional space and time
dynamics are important and can shift substantially, while in the case
of river plumes one can often limit the scope to cover only lateral
spatial elements. However, the methodological framework can be
extended to higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity dependent on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources, one can augment information from synthetic ocean models,
remote-sensing data, or fixed-location sensors on buoys. Traditional
operations and surveys with AUVs are usually limited to observations
along fixed transects on a spatial grid, pre-programmed by the human
operator. A more effective approach is to instead use onboard algorithms to continuously evaluate,
update, and refine future sampling locations (sense-plan-act cycle in Fig. \ref{fig:envir1}), making the information
gathering \emph{adaptive} and \emph{contextual} to the survey
requirements \citep{das11b,fossum18b,fossuminformation}. The space of
sampling opportunities can still be limited to a grid, but adaptation
allows more flexibility. For instance, an AUV could reconstruct or
modify a survey line based on what temperatures it is measuring by
using embedded methods in decision-theoretic planning
\citep{py10,Rajan12,Rajan12b}.

@@

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop. Myopic sampling, i.e. step-wise selection of
the path (on the graph) which reduces the information criterion the
most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward the
designs that improve the classification of water masses as a means to
characterize the frontal zone and thereby the extent of the river plume.

@@

We consider a lateral domain near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes:
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $g/kg$) at location $\bx$. The generalization
to more than two variables is given in Section \ref{gen_k}, and the
adaptive sampling strategies are described in Section
\ref{sec:heuristics}.

@@

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}) and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense,
our work is similar to \cite{bect2012}, \cite{chevalier2014fast}, and
\cite{azzimonti2016quantifying} who describe analytical results for
the IBV for univariate processes. A major contribution of our work is
to derive closed-form results for this design criteria for situations
where the underlying model is based on multivariate GPs.

@@

We discretize the spatial domain $\mathcal{M}$ to a set of $n$ grid
locations $\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell
has area $\Delta$; the grid is used for the waypoint graph setting the
possible design locations. It is also used when solving integral
expressions in Eq. (\ref{measVAR}) and (\ref{sur}), which are then
approximated by sums over all grid cells. Using vector notation, we
denote the Gaussian temperature and salinity variables on the grid and
its density function by the formula:
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma). %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

An important concept underlying this work is the potential of robotic
sampling to adapt survey plans based on data and statistical methods
in accordance with the sense-plan-act control loop (Fig. \ref{fig:envir1}). We
will therefore present a few adaptive strategies for robotic sampling
that build on the notion of multivariate ES.

@@

An adaptive AUV survey is split into many rounds (stages). The selection --on where to sample-- is made on the graph and defined by the nearest grid nodes in the domain $\mathcal{M}_g$.

The optimal sequential design not only considers the best current AUV
grid node, but also how the data gathered at this node could inform
future sampling at successive stages. Let $d^{j,s}$ denote design
number $j$ at stage $s$ of a sequential survey. If this design is
selected, data $\by^{j,s}$ will be gathered. The optimal path
selection situation is depicted in Fig. \ref{fig:PathSelOpt},
\begin{figure}[b!]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a
series of intermixed maximisations over designs and integrals over
data. In practice, the optimal solution is intractable because of
the enormous growth over stages (see
e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}).
Instead, we outline heuristic strategies. In terms of notation, let
$\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote
the data gathered until stage $s-1$, using a selected design
$d^{j,r}$, $r=1,\ldots,s-1$.

@@

\subsection{Myopic Path Planning}
\label{sec:myopic}

@@

\subsection{Look-ahead Trajectory Planning}
\label{sec:LA}

@@

We use a bivariate GP model for temperature and salinity, where we
specify the prior mean:
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{T,0} + \beta_{T,1} x_{\mbox{West}} \\
\beta_{T,1} + \beta_{S,1} x_{\mbox{West}}
\end{bmatrix}.
\end{equation}
This situation
mimics that of a river mouth opening from the south, with the water
masses pulled to the east. The regression parameters are  $\beta_{T,1}=0.065$ and
$\beta_{S,1}=0.1$ for the slope, with west coordinate and $\beta_{T,0}=5.8$
and $\beta_{S,1}=29.0$ for the intercept at a reference map coordinate.

@@

We compare three different static designs denoted
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} with the three described sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. The static sampling paths are pre-determined and cannot be altered and represent the pre-planned strategies used in most current AUV
operational survey designs.

@@

The tables below explore the sensitivity of the result to different
prior correlation values between temperature measured onboard the AUV,
temperature and salinity, prior standard deviations, and spatial
correlations (see table rows). We only compare results at the end of
each survey. The expected ES variances are shown in Table
\ref{tab:sim_res_ev}, results of RMSE are in Table
\ref{tab:sim_res_rmse}, and explained variance is in Table
\ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and
\textit{look-ahead} strategies perform the best in terms of expected
variance in the ES. Surprisingly, the \textit{myopic} strategy
performs slightly better than the \textit{look-ahead}. Part of the
reason for this discrepancy can be traced to the act of pruning some
of the paths for the \textit{look-ahead} strategy; however, even with
less pruning, the \textit{look-ahead} does not outperform the
\textit{myopic} approach. The \textit{static\_north} design is clearly
better than the other two static designs, the reason being that the
others do not focus on the areas where the ES is really uncertain
(centre locations are the most uncertain). By measuring temperature
alone, the expected variances in the ES are only slightly higher than
the combined temperature-salinity estimate. This occurs because the
correlation of $0.6$ means that temperature observations provide
information about salinity as well. For the other predictive measures,
\textit{myopic} and \textit{look-ahead} strategies do well, clearly
beating the \textit{naive} strategy.

@@

\subsection{Model Specification}
\label{sec:exp_modeling}
The statistical model parameters were specified based on a short
preliminary survey where the AUV made an initial transect to determine
environmental conditions and correlation structure. Based on the
initial data, the trend parameters were estimated by linear
regression, where both temperature and salinity are assumed to
increase linearly, going west from the river mouth. Next, the
residuals from the regression analysis were analyzed to study the fit
of the GP model and to specify the covariance parameters.

@@

Fig. \ref{fig:parest} shows diagnostic plots of this
analysis. Fig. \ref{fig:parest_a} shows a cross-plot of temperature
and salinity residuals after the westerly trends in both salinity and
temperature are subtracted from the data. This scatter-plot of joint
residuals indicates larger variability in salinity than in
temperature, and a positive correlation ($0.5$) between the two
variables. Based on the fitted bivariate Gaussian model (ellipses in
Fig. \ref{fig:parest_a}), we can compute the modeled quadratic form of
the residuals, and if the model is adequate they should be
approximately $\chi^2_2$ distributed. Fig. \ref{fig:parest_b} shows
the empirical cumulative distribution function (CDF) of these
quadratic forms (solid) together with the theoretical CDF of the
$\chi^2_2$ distribution. The modeled and theoretical curves are very
similar, which indicates that the Gaussian model fits reasonably
well. Even though there appears to be some clustering in both
Fig. \ref{fig:parest_a} and \ref{fig:parest_b}, the bivariate
diagnostic plots look reasonable and justify a Gaussian
model. Fig. \ref{fig:parest_c} shows the empirical variogram of the
scaled residuals for temperature and salinity. The decay is similar
for the two, and seems to be negligible after about $150$ m.

@@

The sampling locations were distributed over an equilateral grid, as
shown in the grey-colored lattice in Fig. \ref{fig:map}. The robotic
platform consisted of a Light AUV \citep{sousa2012lauv}
(Fig. \ref{fig:lauv}) equipped with a 16 Hz Seabird Fastcat-49
conductivity, temperature, and depth (CTD) sensor providing
temperature and salinity
measurements. %The accuracy of the CTD instrument is $\pm 0.0003$ S/m (conductivity) and $\pm0.002\,^{\circ}\mathrm{C}$ (temperature).
The sampling agent was built on top of the autonomous agent framework
Teleo-Reactive EXecutive (\textit{T-REX})
\citep{py10,Rajan12,Rajan12b}, running an instance of the
\textit{myopic} strategy from Section \ref{sec:myopic} to
control the AUV and decide between sampling locations.

@@

The sampling strategy was designed around the concept of visiting
waypoints sequentially. Arriving at a desired waypoint with new
measurements and an updated model, the AUV triggers the myopic
strategy to evaluate the different design criteria (see
Eq. \eqref{critSEQ}). The waypoint-and-path combination that is
expected to reduce the IBV the most is selected, and upon arrival this
procedure is then repeated. At each stage, it takes the AUV about 30
seconds to evaluate the expected IBV for all the possible
waypoint-and-path alternatives.

The AUV was set to start in the south-center part of the waypoint
graph, with the previously outlined GP model of the environment (Section \ref{sec:exp_modeling}). A
survey was set to take approximately 40 minutes, visiting 15 waypoints
on the grid, with the AUV running near the surface to capture the
plume. On its path from one waypoint to the next, the AUV gathered
data regularly, and the GP model assimilated temperature and salinity
data with an update frequency of 30 seconds, giving about three updates
per stage.
